---
title: "Preparing FASTA and annotation files for RiboViz"
output:
  html_notebook: default
  html_document: default
---

### Packages
```{r, eval=FALSE}
library(Biostrings)
library(rtracklayer)
library(GenomicRanges)
library(parallel)
```

# Yeast datasets

### Prepare S288c annotations and ORF seqs
```{r, eval=FALSE}
# Download annotations and genome seq from here - http://downloads.yeastgenome.org/sequence/S288C_reference/genome_releases/S288C_reference_genome_R64-2-1_20150113.tgz
yeast_annot <- readGFFAsGRanges("./saccharomyces_cerevisiae_R64-2-1_20150113.gff")
yeast_genome <- readDNAStringSet("./S288C_reference_sequence_R64-2-1_20150113.fsa")

# Pad genomes with 250 N nts to provide null flanking regions for genes close to chromosomal boundaries
yeast_genome <- xscat(DNAString(paste(rep("N",250),collapse="")),yeast_genome,DNAString(paste(rep("N",250),collapse="")))
names(yeast_genome) <- paste("chr",c(as.character(as.roman(1:16)),"mt"),sep="")

# Adjust the annotations for this 250nt shift
yeast_annot <- shift(yeast_annot,250)

# Get the annotations for all CDS of uncharacterized and verified genes
cds_annot <- yeast_annot[(yeast_annot$orf_classification=="Uncharacterized" | yeast_annot$orf_classification=="Verified") & !is.na(yeast_annot$orf_classification) & yeast_annot$type=="CDS"]
cds_annot$Name <- sapply(strsplit(cds_annot$Name,"_"), `[`, 1)

```

### Get ORF+UTR seqs
```{r, eval=FALSE}
# Funciton to resize GRanges while keeping track of multi-exon genes
resize_gff_flank <- function(gene_location,flank){
  if(length(gene_location)==1){	
    start(gene_location) <- start(gene_location)-flank
    end(gene_location) <- end(gene_location)+flank
  }else{
    start(gene_location)[start(gene_location)==min(start(gene_location))] <- start(gene_location)[start(gene_location)==min(start(gene_location))]-flank
    end(gene_location)[end(gene_location)==max(end(gene_location))] <- end(gene_location)[end(gene_location)==max(end(gene_location))]+flank
  }
  return(gene_location)
}

# Resize all CDS with 250 nt UTR regions
tmp_gff <- sapply(unique(cds_annot$Name),function(x){gene_location <- cds_annot[cds_annot$Name==x]; resize_gff_flank(gene_location = gene_location, flank=250)})
cds_flank_annot <- unlist(GRangesList(tmp_gff))

# Obtain sequences of CDS+UTRs from genome
cds_flank_seq <- yeast_genome[cds_flank_annot]
names(cds_flank_seq) <- cds_flank_annot$Name

# Combine multi-exon genes into a single transcript
output_seqlist <- list()
cc <- 1
for(i in unique(names(cds_flank_seq))){
  output_seqlist[[cc]] <- c(unlist(cds_flank_seq[names(cds_flank_seq)==i]))
  cc <- cc+1
}
output_seq <- DNAStringSet(output_seqlist)
names(output_seq) <- unique(names(cds_flank_seq))

# Write seqs to a FASTA file
writeXStringSet(output_seq,filepath = "yeast_CDS_w_250utrs.fa",format = "fasta")
```

### Create a GFF file based on these transcripts
```{r, eval=FALSE}
len_gff <- c(matrix(c(rep(250,length(output_seq)), (width(output_seq)-500), rep(250,length(output_seq))),nrow=3,byrow = T))
start_gff <- c(matrix(c(rep(1,length(output_seq)),rep(251,length(output_seq)), (width(output_seq)-249)),nrow=3,byrow = T))
type_gff <- rep(c("UTR5","CDS","UTR3"),length(output_seq))

# Create a GRange object to be saved as a GFF
output_gff <- GRanges(seqnames = rep(names(output_seq),each=3), ranges = IRanges(start=start_gff, width=len_gff),strand = "+",type=type_gff,Name=rep(names(output_seq),each=3))
export.gff3(output_gff, con=file("yeast_CDS_w_250utrs.gff3","w"))
```

### Create an RData object with gene-specific codon positions for riboviz plots
```{r, eval=FALSE}
seq <- output_seq
gff <- output_gff

seq <- seq[gff[gff$type=="CDS"]] # Restrict sequences to only CDS
seq <- seq[width(seq)>600] # Remove sequences <200 codons

seq <- DNAStringSet(seq,start=601) # Trim the first 200 codons from each CDS
cods <- sapply(sapply(seq,codons),as.character) # Split the sequences into codons

# Function to find positions for a codon in a CDS
codon_pos_locator <- function(seq,codon,gene){
  out <- NULL
  pos <- which(seq==codon)
  if(length(pos)>=1)
    out <- cbind(gene,pos)
  return(out)
}

out <- sapply(names(GENETIC_CODE),
              function(codon){
                allpos <- mclapply(names(cods),
                                function(gene){
                                  codon_pos_locator(seq=cods[[gene]],
                                                    codon=codon,
                                                    gene=gene)
                                },mc.cores=50)
                do.call(rbind,allpos)
              })

# Save output as an RData object
save(out, "yeast_codon_pos_i200.RData")
```

# Human datasets

### Prepare GRCh38 annotations
```{r, eval=FALSE}
# Download annotations from here - https://www.gencodegenes.org/releases/21.html
hsap_annot <- readGFFAsGRanges("gencode.v21.annotation.gff3.gz")

# Create a new transcript id column without versions ()
mcols(hsap_annot)$tid <- sapply(strsplit(hsap_annot$transcript_id,"\\."), `[`, 1)
```

### Find the principal transcript for each gene
```{r, eval=FALSE}
# Download appris annotations from here - http://appris.bioinfo.cnio.es/

# Read in Appris list
appris <- read.table("human_appris_data.principal.txt")

# Get a list of all principal transcripts
princ_app <- appris[grep("P",as.character(appris[,5])),]
princ_app <- princ_app[order(princ_app$V1),]

# Pick the first principal transcript in genes with multiple principal transcripts
all_enstt <- princ_app[!duplicated(princ_app[,1]),]
```

### Subset from official cdna list
```{r, eval=FALSE}
# Download CDNA seqs from here - http://ftp.ensembl.org/pub/release-83/fasta/homo_sapiens/cdna/

# Read in CDNA sequences
cdna_all <- readDNAStringSet("Homo_sapiens.GRCh38.cdna.all.fa.gz")
cdna_names <- sapply(strsplit(names(cdna_all)," "), `[`, 1)
cdna_names2 <- sapply(strsplit(cdna_names,"\\."), `[`, 1)

# Subset principal appris CDNAs and rename the transcripts 
cdna_sub <- cdna_all[cdna_names2 %in% all_enstt$V3]
names(cdna_sub) <- cdna_names2[cdna_names2 %in% all_enstt$V3]

# Exon annotations
exon_annot <- hsap_annot[hsap_annot$type=="exon"]

# Get transcript lengths from exon annotations
trans_length <- mclapply(names(cdna_sub), function(x){
  sum(width(exon_annot[exon_annot$tid==x]))
  },mc.cores = 50)

# Subset appris CDNAs whose lengths match with the annotations that are at least 150 bp
cdna_sub <- cdna_sub[width(cdna_sub)==trans_length & width(cdna_sub)>=150]

# Get CDS annotations for transcripts 
cds_annot <- hsap_annot[hsap_annot$type=="CDS" & hsap_annot$tid %in% names(cdna_sub) ]
exon_annot <- exon_annot[exon_annot$tid %in% unique(cds_annot$tid)]

# Subset appris CDNAs that are CDS
cdna_sub <- cdna_sub[names(cdna_sub) %in% cds_annot$tid]

# Save the appris cdna seqs as a fasta file
writeXStringSet(cdna_sub,filepath = "human_appris_cdna.fa",format = "fasta")
```

### Create a GFF file based on these transcripts
```{r, eval=FALSE}
# Function to get UTR and CDS lengths in each transcript
get_utr_cds_len <- function(exon,cds,strand){
  posdiff <- unlist(setdiff(ranges(exon),ranges(cds)))
  utr5 <- sum(posdiff<min(start(cds)))
  clen <- sum(width(cds))
  utr3 <- sum(posdiff>max(end(cds)))
  if(strand=="-"){
    tmp <- utr3
    utr3 <- utr5
    utr5 <- tmp
  }
  return(c(utr5,clen,utr3))
}

# Get all UTR and CDS lengths
all_lengths <- mclapply(names(cdna_sub),function(x){
  exon <- exon_annot[exon_annot$tid==x]
  cds <- cds_annot[cds_annot$tid==x]
  strand <- as.character(strand(cds))[1]
  get_utr_cds_len(exon = exon, cds = cds, strand = strand)
}, mc.cores = 50)

# Convert the output to a matrix such that the 3 columns represent utr5, cds, utr3
length_mat <- matrix(unlist(all_lengths), ncol=3, byrow=T)

len_gff <- c(t(length_mat))
start_gff <- c(matrix(c(rep(1,nrow(length_mat)),(length_mat[,1]+1), (length_mat[,1]+length_mat[,2]+1)),nrow=3,byrow = T))
type_gff <- rep(c("UTR5","CDS","UTR3"),nrow(length_mat))

# Create a GRange object to be saved as a GFF
output_gff <- GRanges(seqnames = rep(names(cdna_sub),each=3), ranges = IRanges(start=start_gff, width=len_gff),strand = "+",type=type_gff,Name=rep(names(cdna_sub),each=3))
export.gff3(output_gff, file.path("human_appris_cdna.gff3"))
```

### Create an RData object with gene-specific codon positions for riboviz plots
```{r, eval=FALSE}
seq <- output_seq
gff <- output_gff

seq <- seq[gff[gff$type=="CDS"]] # Restrict sequences to only CDS
seq <- seq[width(seq)>600] # Remove sequences <200 codons

seq <- DNAStringSet(seq,start=601) # Trim the first 200 codons from each CDS
seq <- seq[width(seq)%%3==0] # Ignore any transcripts with frame-shifts
cods <- sapply(sapply(seq,codons),as.character) # Split the sequences into codons

# Function to find positions for a codon in a CDS
codon_pos_locator <- function(seq,codon,gene){
  out <- NULL
  pos <- which(seq==codon)
  if(length(pos)>=1)
    out <- cbind(gene,pos)
  return(out)
}

out <- sapply(names(GENETIC_CODE),
              function(codon){
                allpos <- mclapply(names(cods),
                                function(gene){
                                  codon_pos_locator(seq=cods[[gene]],
                                                    codon=codon,
                                                    gene=gene)
                                },mc.cores=50)
                do.call(rbind,allpos)
              })

# Save output as an RData object
save(out, "human_codon_pos_i200.RData")
```

# Mouse datasets

### Prepare GRCh38 annotations
```{r, eval=FALSE}
# Download annotations from here - https://www.gencodegenes.org/releases/21.html
mouse_annot <- readGFFAsGRanges("./gencode.vM14.annotation.gff3.gz")

# Create a new transcript id column without versions ()
mcols(mouse_annot)$tid <- sapply(strsplit(mouse_annot$transcript_id,"\\."), `[`, 1)
```

### Find the principal transcript for each gene
```{r, eval=FALSE}
# Download appris annotations from here - http://appris.bioinfo.cnio.es/

# Read in Appris list
appris <- read.table("mouse_appris_data.principal.txt")

# Get a list of all principal transcripts
princ_app <- appris[grep("P",as.character(appris[,5])),]
princ_app <- princ_app[order(princ_app$V1),]

# Pick the first principal transcript in genes with multiple principal transcripts
all_enstt <- princ_app[!duplicated(princ_app[,1]),]
```

### Subset from official cdna list
```{r, eval=FALSE}
# Download CDNA seqs from here - http://ftp.ensembl.org/pub/release-83/fasta/mus_musculus/cdna/

# Read in CDNA sequences
cdna_all <- readDNAStringSet("./Mus_musculus.GRCm38.cdna.all.fa.gz")
cdna_names <- sapply(strsplit(names(cdna_all)," "), `[`, 1)
cdna_names2 <- sapply(strsplit(cdna_names,"\\."), `[`, 1)

# Subset principal appris CDNAs and rename the transcripts 
cdna_sub <- cdna_all[cdna_names2 %in% all_enstt$V3]
names(cdna_sub) <- cdna_names2[cdna_names2 %in% all_enstt$V3]

# Exon annotations
exon_annot <- mouse_annot[mouse_annot$type=="exon"]

# Get transcript lengths from exon annotations
trans_length <- mclapply(names(cdna_sub), function(x){
  sum(width(exon_annot[exon_annot$tid==x]))
  },mc.cores = 50)

# Subset appris CDNAs whose lengths match with the annotations that are at least 150 bp
cdna_sub <- cdna_sub[width(cdna_sub)==trans_length & width(cdna_sub)>=150]

# Get CDS annotations for transcripts 
cds_annot <- mouse_annot[mouse_annot$type=="CDS" & mouse_annot$tid %in% names(cdna_sub) ]
exon_annot <- exon_annot[exon_annot$tid %in% unique(cds_annot$tid)]

# Subset appris CDNAs that are CDS
cdna_sub <- cdna_sub[names(cdna_sub) %in% cds_annot$tid]

# Save the appris cdna seqs as a fasta file
writeXStringSet(cdna_sub,filepath = "mouse_appris_cdna.fa",format = "fasta")
```

### Create a GFF file based on these transcripts
```{r, eval=FALSE}
# Function to get UTR and CDS lengths in each transcript
get_utr_cds_len <- function(exon,cds,strand){
  posdiff <- unlist(setdiff(ranges(exon),ranges(cds)))
  utr5 <- sum(posdiff<min(start(cds)))
  clen <- sum(width(cds))
  utr3 <- sum(posdiff>max(end(cds)))
  if(strand=="-"){
    tmp <- utr3
    utr3 <- utr5
    utr5 <- tmp
  }
  return(c(utr5,clen,utr3))
}

# Get all UTR and CDS lengths
all_lengths <- mclapply(names(cdna_sub),function(x){
  exon <- exon_annot[exon_annot$tid==x]
  cds <- cds_annot[cds_annot$tid==x]
  strand <- as.character(strand(cds))[1]
  get_utr_cds_len(exon = exon, cds = cds, strand = strand)
}, mc.cores = 50)

# Convert the output to a matrix such that the 3 columns represent utr5, cds, utr3
length_mat <- matrix(unlist(all_lengths), ncol=3, byrow=T)

len_gff <- c(t(length_mat))
start_gff <- c(matrix(c(rep(1,nrow(length_mat)),(length_mat[,1]+1), (length_mat[,1]+length_mat[,2]+1)),nrow=3,byrow = T))
type_gff <- rep(c("UTR5","CDS","UTR3"),nrow(length_mat))

# Create a GRange object to be saved as a GFF
output_gff <- GRanges(seqnames = rep(names(cdna_sub),each=3), ranges = IRanges(start=start_gff, width=len_gff),strand = "+",type=type_gff,Name=rep(names(cdna_sub),each=3))
export.gff3(output_gff, file.path("mouse_appris_cdna.gff3"))
```

### Create an RData object with gene-specific codon positions for riboviz plots
```{r, eval=FALSE}
seq <- output_seq
gff <- output_gff

seq <- seq[gff[gff$type=="CDS"]] # Restrict sequences to only CDS
seq <- seq[width(seq)>600] # Remove sequences <200 codons

seq <- DNAStringSet(seq,start=601) # Trim the first 200 codons from each CDS
seq <- seq[width(seq)%%3==0] # Ignore any transcripts with frame-shifts
cods <- sapply(sapply(seq,codons),as.character) # Split the sequences into codons

# Function to find positions for a codon in a CDS
codon_pos_locator <- function(seq,codon,gene){
  out <- NULL
  pos <- which(seq==codon)
  if(length(pos)>=1)
    out <- cbind(gene,pos)
  return(out)
}

out <- sapply(names(GENETIC_CODE),
              function(codon){
                allpos <- mclapply(names(cods),
                                function(gene){
                                  codon_pos_locator(seq=cods[[gene]],
                                                    codon=codon,
                                                    gene=gene)
                                },mc.cores=50)
                do.call(rbind,allpos)
              })

# Save output as an RData object
save(out, "mouse_codon_pos_i200.RData")
```